Approximation as Differentiation

### 2025-05-08 main/formal-structures.tex
Move semi-additive structure on morphisms para to beginning of section on semi-additive categories.

As sketched in \secref{introduction:galois-slicing}, Galois slicing requires the approximants of a point $x$
to form a bounded lattice $(X, \meet, \join, \top, \bot)$, with least element $\bot$ representing the
approximation that discards all information about $x$, greatest element $\top$ representing the approximation
that retains all information about $x$, and $\meet$ and $\join$ providing two canonical ways to combine
approximations. Thus rather than working in $\PosGal$, we define the following subcategory.

@@

\subsubsection{Additive structure on morphisms.}

A bounded semilattice is an example of a \emph{commutative monoid} $(X, +, 0)$, namely a set $X$ equipped with
an associative, commutative binary operation $+$ with unit $0$. One observation about commutative monoids is
that pointwise lifting of $+$ to homomorphisms $f, g: X \to Y$, where $(f + g)(x) = f(x) + g(x): X \to Y$, is
also ``additive'', in the sense of being associative and commutative and having a ``zero'' homomorphism
$0_{X,Y}: X \to Y$ given by the constant map sending any $x \in X$ to $0_Y$, which acts as a unit for addition
of homomorphisms. The homomorphisms from $X$ to $Y$ thus have a commutative monoid structure themselves.

In particular $\LatGal$ has this structure..

% Moreover since $(X, \top, \meet)$ is a commutative monoid, and thus a $\CMon$-enriched (poset)
% category, for any $X, Y$ there is a zero meet-semilattice homomorphism $0^{\meet}_{X,Y} = \const(\top_Y): X
% \to Y$ and also (idempotent) addition of homomorphisms $f +_{\meet} g: X \to Y$ given pointwise by $\meet$.
% Dually, there is a zero join-semilattice homomorphism $0^{\join}_{X,Y} = \const(\bot_Y): X \to Y$ and
% idempotent addition of homomorphisms $f +_{\join} g: X \to Y$ given pointwise by $\join$.

@@

@@

\vspace{-2mm}
\begin{center}
\begin{tikzcd}
   X \arrow[r, "\diag"] & X \biprod X \arrow[r, "f \biprod g"] & Y \biprod Y \arrow[r, "\codiag"] & Y
\end{tikzcd}
\end{center}

@@

% Done:
% (1) introduce the category of lattices and Galois connections, which is what we will use to interpret forward and backward approximation, and establish some properties of this category;

\note{Probably a better way: (2) say that we want sets with each point having an associated lattice of approximations, and morphisms to match; (3) note that this is exactly the Grothendieck construction for a particular indexed category; (4) Say that CHAD is what we need here; (5) Recapitulate the bits of CHAD that we need, with our Agda formalisation; (6) Explain how we cope with higher-order structure, by using presheaves and the Yoneda embedding.}

>>>>

As sketched in \secref{introduction:galois-slicing}, Galois slicing requires that the approximants of a point
$x$ form a bounded lattice $(X, \meet, \join, \top, \bot)$, with least element $\bot$ representing the
approximation that discards all information about $x$, greatest element $\top$ representing the approximation
that retains all information about $x$, and $\meet$ and $\join$ providing two canonical ways to combine
approximations. Thus rather than working directly in $\PosGal$, we define the following subcategory.

@@

@@

An important observation about the semilattice homomorphisms that appear in adjoint pairs in $\LatGal$ is that
pointwise lifting of $\meet$ to meet-semilattice homomorphisms $f, g: X \to Y$, where $(f \meet g)(x) = f(x)
\meet g(x): X \to Y$, gives a notion of ``addition'' for these maps which is associative and commutative, and
which has a ``zero'' homomorphism $0_{X,Y}: X \to Y$ given by the constant map sending any $x \in X$ to $0_Y$,
acting as a unit for addition of homomorphisms. The same can be said for the join-preserving maps from $X$ to
$Y$, which also have a commutative monoid structure; Galois connections inherit this structure component-wise.
We now say more generally how this so-called \emph{semi-additive} structure on morphisms arises and explain
its significance to adapting CHAD to provide a semantic basis for Galois slicing as well as automatic
differentiation.

@@

\begin{center}
\begin{tikzcd}
   X \arrow[r, "\diag"] & X \biprod X \arrow[r, "f \biprod g"] & Y \biprod Y \arrow[r, "\codiag"] & Y
\end{tikzcd}
\end{center}

@@

\note{Probably a better way: (1) introduce the category of lattices and Galois connections, which is what we will use to interpret forward and backward approximation, and establish some properties of this category; (2) say that we want sets with each point having an associated lattice of approximations, and morphisms to match; (3) note that this is exactly the Grothendieck construction for a particular indexed category; (4) Say that CHAD is what we need here; (5) Recapitulate the bits of CHAD that we need, with our Agda formalisation; (6) Explain how we cope with higher-order structure, by using presheaves and the Yoneda embedding.}
